---
title: "Cohort Analysis"
output: 
  md_document:
    variant: gfm
---

```{r echo=FALSE}
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(lubridate)
```

Data can be downloaded from [here](https://www.kaggle.com/gspmoreira/articles-sharing-reading-from-cit-deskdrop)

```{r}
df <- read.csv("~/code/notebooks/article-rc/data/users_interactions.csv")
df$date <- as.Date(as.POSIXct(df$timestamp, origin = "1970-01-01"))
```


```{r}
coh <- data.frame(q = quarter(df$date, with_year = TRUE),
                  user = df$personId,
                  event = df$eventType)

glimpse(coh)
```

```{r}
coh <- coh %>% 
  group_by(user) %>% 
  mutate(first = min(q)) %>% 
  group_by(first, q) %>% 
  summarise(users = n_distinct(user)) %>% 
  spread(q, users)

coh
```

```{r}
shiftrow <- function(v) {
  first_na_index <- min(which(!is.na(v)))
  
  c(v[first_na_index:length(v)], rep(NA, first_na_index - 1))
}

shifted <- data.frame(
  cohort = coh$first,
  t(apply(select(as.data.frame(coh), 2:ncol(coh)), 1, shiftrow))
)

colnames(shifted) <- c("cohort", sub("", "quarter.",
                                     str_pad(1:(ncol(shifted) - 1), 2, pad = "0")))

shifted
```

```{r}
shifted_pct <- data.frame(
  cohort = shifted$cohort,
  shifted[,2:nrow(shifted)] / shifted[["quarter.01"]]
)

shifted_pct
```


```{r}
p_data_abs <- gather(shifted, "cohort_age", "people", 2:ncol(shifted))
p_data_pct <- gather(shifted_pct, "cohort_age", "percent", 2:ncol(shifted_pct))

labelnames <- c(p_data_abs$people[1:(ncol(shifted)-1)],
                p_data_pct$percent[(ncol(shifted)):(nrow(p_data_pct))])

pretty_print <- function(n) {
  label <- case_when(n <= 1 ~ sprintf("%1.0f %%", n*100),
                     n > 1 ~ as.character(n),
                     TRUE ~ " ")
  
  label
}
```

```{r}
p_data <- data.frame(
  cohort = p_data_pct$cohort,
  cohort_age = p_data_pct$cohort_age,
  percentage = p_data_pct$percent,
  label = pretty_print(labelnames)
)
```

```{r}
ggplot(p_data, aes(x = cohort_age, y = reorder(cohort, desc(cohort)))) +
  geom_raster(aes(fill = percentage)) +
  scale_fill_continuous(guide = FALSE) +
  geom_text(aes(label = label), color = "white") +
  labs(x = "cohort_age", y = "cohort",
       title = "Retention table",
       subtitle = "Quarterly breakdown")
```

## Monthly breakdown

```{r}
coh <- data.frame(month = format.Date(df$date, "%Y-%m"),
                  user = df$personId,
                  event = df$eventType,
                  stringsAsFactors = FALSE)

glimpse(coh)
```

```{r}
coh <- coh %>% 
  group_by(user) %>% 
  mutate(first = min(month)) %>% 
  group_by(first, month) %>% 
  summarise(users = n_distinct(user)) %>% 
  spread(month, users)

coh
```

```{r}
shifted <- data.frame(
  cohort = coh$first,
  t(apply(select(as.data.frame(coh), 2:ncol(coh)), 1, shiftrow))
)

colnames(shifted) <- c("cohort", sub("", "month.",
                                     str_pad(1:(ncol(shifted) - 1), 2, pad = "0")))

shifted
```

```{r}
shifted_pct <- data.frame(
  cohort = shifted$cohort,
  shifted[,2:nrow(shifted)] / shifted[["month.01"]]
)

shifted_pct
```

```{r}
p_data_abs <- gather(shifted, "cohort_age", "people", 2:ncol(shifted))
p_data_pct <- gather(shifted_pct, "cohort_age", "percent", 2:ncol(shifted_pct))

labelnames <- c(p_data_abs$people[1:(ncol(shifted)-1)],
                p_data_pct$percent[(ncol(shifted)):(nrow(p_data_pct))])

pretty_print <- function(n) {
  label <- case_when(n <= 1 ~ sprintf("%1.0f %%", n*100),
                     n > 1 ~ as.character(n),
                     TRUE ~ " ")
  
  label
}
```

```{r}
p_data <- data.frame(
  cohort = p_data_pct$cohort,
  cohort_age = p_data_pct$cohort_age,
  percentage = p_data_pct$percent,
  label = pretty_print(labelnames)
)
```

```{r}
ggplot(p_data, aes(x = cohort_age, y = reorder(cohort, desc(cohort)))) +
  geom_raster(aes(fill = percentage)) +
  scale_fill_continuous(guide = FALSE) +
  geom_text(aes(label = label), color = "white") +
  labs(x = "cohort_age", y = "cohort",
       title = "Retention Table",
       subtitle = "Monthly breakdown")
```

From the plot above, it is clear that the retailer has issues on 2 areas:  

* Acquisition (cohort size worsening over time)
* Retention (same as above for first month retention)

It would be interesting to understand (might be added later) which `eventType` has
suffered the most over time to figure where marketing effort should be focused.



